---
title: "Measles in San Francisco"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
```

Measles in San Francisco is interesting. What's going on?

```{r}
library(kernlab)
library(mgcv)
library(dplyr)
library(scam)
library(tsiR)
library(glmmTMB)
library(ggplot2); theme_set(theme_bw())
source("../R/reconstruct.R")
source("../R/simulate.R")
```

First, let's look at the data:

```{r}
measles_data <- read.csv("measlesUKUS.csv")

measles_US <- measles_data %>% 
	filter(country=="US") %>%
	mutate(cases=ifelse(is.na(cases), 0, cases))

measles_list <- measles_US %>%
	split(as.character(.$loc))

sf <- measles_list$`SAN FRANCISCO`

plot(sf$decimalYear, sf$cases, type="l",
	 ylab="Biweekly cases",
	 xlab="time (year)")
```

Periodogram suggests that there are strong three year and one year cycles.

```{r}
ss <- spectrum(sf$cases, plot=FALSE, log="no")

plot(1/ss$freq/26, ss$spec, type="l", xlim=c(0, 6),
	 ylab="spectrum",
	 xlab="period (years)")
abline(v=3, lty=2)
```

We want to look at epidemic from 1921 to 1940 (1920 has a weird reporting problem..):

```{r}
sf2 <- sf %>% 
	filter(year > 1920, year <= 1940)
```

Compare susceptible dynamics (quite sensitive to time period):

```{r}
rr_scam <- reconstruct_scam(sf$cases, sf$rec)
rr_gauss <- reconstruct_gauss(sf$cases, sf$rec)
rr_scam2 <- reconstruct_scam(sf2$cases, sf2$rec)
rr_gauss2 <- reconstruct_gauss(sf2$cases, sf2$rec)

plot(sf$decimalYear, rr_scam$Z, type="l", xlim=c(1920, 1940), lwd=2)
lines(sf2$decimalYear, rr_scam2$Z, col=1, lty=2, lwd=2)
lines(sf$decimalYear, rr_gauss$Z, col=2, lwd=2)
lines(sf2$decimalYear, rr_gauss2$Z, col=2, lty=2, lwd=2)
```

Estimated reporting rate is also quite variable

```{r}
plot(sf$decimalYear, 1/rr_scam$rho, type="l", ylim=c(0, 1), lwd=2, xlim=c(1920, 1940))
lines(sf2$decimalYear, 1/rr_scam2$rho, col=1, lwd=2, lty=2)
lines(sf$decimalYear, 1/rr_gauss$rho, col=2, lwd=2)
lines(sf2$decimalYear, 1/rr_gauss2$rho, col=2, lwd=2, lty=2)
```

For now, we are going to use scam2 results. scam appears to regularize things a little better than gausspr...

```{r}
Iprev <- head(sf2$cases * rr_scam2$rho, -1)
Inew <- tail(sf2$cases * rr_scam2$rho, -1)

plot(Iprev, Inew)
```

Let's try to fit the TSIR model assuming Poisson:

```{r}
fitdata <- data.frame(
	time=sf2$decimalYear,
	cases=sf2$cases,
	pop=sf2$pop,
	births=sf2$rec
)

ff <- runtsir(fitdata,
			  regtype = "user",
			  userYhat=unname(c(rr_scam2$Yhat)),
			  family="poisson",
			  link="log",
			  method="negbin")
```

What happens?

```{r}
plotres(ff)
```

The simulation predicts a consist annual cycle.
What can we do to fix this? First, let's look at the dispersion patterns:

```{r}
plot(exp(predict(ff$glmfit)), (ff$glmfit$data$Inew-predict(ff$glmfit))^2,
	 xlab="Mean",
	 ylab="Variance")
```

Surprisingly, there's a strong quadratic mean-variance relationship (often, it's not this clear). So we need to fit negative binomial instead of Poisson. 

```{r}
ff2 <- runtsir(fitdata,
			  regtype = "user",
			  userYhat=unname(c(rr_scam2$Yhat)),
			  family="nbinom",
			  link="log",
			  method="negbin",
			  Smean=seq(0.03 * min(sf$pop), 0.2 * min(sf$pop), length.out=50))
```

Does this perform any better?

```{r}
plotres(ff2)
```

Not so much. What about dispersion?

```{r}
plot(exp(predict(ff2$glmfit)), (ff$glmfit$data$Inew-predict(ff2$glmfit))^2,
	 xlab="Mean",
	 ylab="Variance")
curve(x * (1 + x/ff2$glmfit$theta), add=TRUE, col=2)
```

What can we do to make this fit better? We can try to simulate with extra variability.

```{r}
set.seed(101)
sim <- simulate.sir(
	betamat=matrix(ff2$contact[,2] * mean(ff2$pop), nrow=1),
	alpha=ff2$alpha,
	mixmatmatrix(1),
	I0=ff2$res[1,1],
	S0=ff2$simS[1,1],
	nsim=50,
	method="nbinom",
	popmat=matrix(head(ff2$pop,-1), nrow=1),
	birthmat=matrix(head(sf2$rec,-1), nrow=1),
	rhomat=matrix(rr_scam2$rho, nrow=1),
	phi=ff2$glmfit$theta
)

simcases <- sim %>%
	sapply("[[", "C")

matplot(simcases, type="l", col=1)
lines(sf2$cases, col=2, lwd=2)
```

We get a lot more interesting dynamics, which may better resemble the dynamics observed in San Francisco.


